\chapter{INTRODUCTION}

{\baselineskip=2\baselineskip
Agriculture has historically depended on traditional practices and empirical knowledge for crop management, harvest timing, and quality assessment. These conventional methods, while functional, are inherently subjective and lack the standardization necessary for consistent and reliable evaluation, particularly in the context of fruit maturity where internal quality attributes are not directly observable \citep{savako_2025_nondestructive}. The increasing adoption of precision agriculture and data-driven sensing technologies represents a significant departure from these practices, offering objective and reproducible alternatives that are more suitable for modern agricultural quality assessment.

A critical consideration in pre-harvest management is the physiological classification of fruit crops. Non-climacteric fruits, including watermelon, citrus, grapes, and pineapple, do not undergo post-harvest ripening and therefore must attain full maturity prior to detachment from the plant. Unlike climacteric varieties such as banana and mango, which continue to ripen after harvest through ethylene-mediated processes \citep{liu_2023_research}, non-climacteric fruits exhibit no further development in sugar content, texture, or flavor once harvested. This biological characteristic makes accurate pre-harvest ripeness assessment a technical requirement rather than an optional quality measure, as premature harvest results in permanent degradation of fruit quality.

Despite this, smallholder farmers in many agricultural regions, including the Philippines, continue to rely on subjective sensory indicators such as sound resonance, surface appearance, and tactile response to determine harvest readiness. These experience-based approaches are inconsistent across individuals, lack quantifiable metrics, and frequently result in the collection of suboptimal produce. The absence of accessible and field-deployable instrumentation for objective ripeness assessment represents a technological gap that this study seeks to address.

Ripeness assessment methods can generally be divided based on two things: whether they damage the fruit, and whether they require physical contact with it. Destructive methods, like cutting or pulping the watermelon to check its internal state, give accurate results but make the fruit unsellable afterward, which is not practical for farmers or vendors. Non-destructive methods avoid this problem by keeping the fruit intact after testing. But within non-destructive methods, there is still a distinction between invasive and non-invasive approaches. Non-invasive methods do not breach or penetrate the fruit at all, such as using imaging or color analysis from a distance, or using thumping methods. Invasive non-destructive methods, on the other hand, breach the interior of the fruit, such as with using probes, but do not cause any lasting damage. The system proposed in this study falls under the non-destructive and non-invasive category, as the watermelon is not damaged in any way during the assessment. 

This chapter presents the background and motivation behind the development of a portable, non-destructive watermelon ripeness detection system that integrates acoustic signal processing and machine learning. It identifies the research problem arising from the limitations of existing manual and technological assessment methods, defines the objectives of the study, and establishes the scope, significance, and limitations within which the proposed system is developed and evaluated.



\section{Background of the Study}

Advanced technologies have become a cornerstone of modern agriculture, streamlining production and quality control through data-driven methods. Recent trends emphasize the integration of machine learning to enhance efficiency through precision farming and improved resource allocation \citep{boterovalencia_2025_machine}. Research indicates a paradigm shift where machine learning models have been useful in recent systems dominating over food production. These approaches allow for the evaluation of internal fruit characteristics without compromising physical integrity, aligning practices with contemporary food safety and environmental management requirements \citep{arajo_2023_machine}.

The shift toward quantitative analysis is further supported by research into automated grading systems. Studies highlight how machine learning algorithms are transforming traditional agricultural practices by providing high-precision sorting and quality assurance \citep{boterovalencia_2025_machine}. Furthermore, the integration of computer vision and acoustic analysis has proven effective in agricultural produce handling, with specialized algorithms enabling accurate and efficient solutions for produce quality management \citep{gowda_2025_automating}. These trends illustrate a growing reliance on feature extraction techniques to interpret complex biological data.

In 2023, the Philippines ranked \#52 in the global production of watermelon, having produced around 124,737.55 metric tons of watermelon (Trends of Fresh Watermelon Production in Philippines, 2024). Although the Philippines’ contribution to the total global production of watermelon is only around 0.12\%, it is still a high-value crop in the Philippines.

Despite these global advancements, many local farmers in the Philippines still rely heavily on traditional, manual methods for pre-harvest assessment. Currently, determining the maturity of watermelons in the field depends on the thumping technique—manually tapping the fruit and listening to the resulting sound, in addition to visual inspection and the observation of the wilting of the watermelon’s tendrils. This practice is subjective, as the interpretation of acoustic resonance varies significantly among individuals. Recent literature notes that subjective harvest decisions can lead to the collection of underripe produce, which lacks the necessary sugar content and flavor profiles required for market standards \citep{savako_2025_nondestructive}. This reliance on human intuition creates an inconsistency in fruit quality that can affect the overall reliability of the harvest.

Acoustic-based methods have gained particular attention due to their low cost, portability, and ease of field implementation, yet their performance remains highly sensitive to environmental noise, with wind, machinery, and handling interference significantly degrading signal quality \citep{pamungkas_2021_evaluation}, and standard FFT-based feature extraction has been shown to be insufficient for capturing the subtle acoustic differences caused by internal ripeness changes \citep{chen_2021_nondestructive}. Similarly, computer vision and optical-based methods are either limited to surface-level external assessment and cannot directly measure internal quality attributes such as flesh firmness or sugar content \citep{defang_2024_intelligent, perezudell_2023_an}, or depend on expensive stationary equipment incompatible with portable farm-level deployment \citep{shah_2025_a, qi_2024_application}.

To overcome these inconsistencies, various nondestructive testing techniques have been explored, including acoustic-based methods such as FFT-based frequency analysis of tapping-induced vibration signals \citep{wongkalasin_2021_watermelon}, FFT magnitude analysis of thumping sounds using a custom mechanical apparatus \citep{okafor_2020_design}, acoustic impulse response measurement using mechanical tapping devices \citep{pamungkas_2021_evaluation}, and wavelet-based acoustic signal decomposition for non-destructive fruit screening \citep{chen_2021_nondestructive}, as well as computer vision-based methods such as HSV color space classification from field photographs \citep{pamungkas_2021_evaluation}, lightweight YOLO-based deep learning detection for real-time ripeness classification \citep{defang_2024_intelligent}, and hyperspectral imaging coupled with deep learning for non-destructive quality estimation \citep{shah_2025_a, qi_2024_application}.

This study proposes a portable system for detecting watermelon ripeness using acoustic signal processing and machine learning. The device will utilize a multi-modal approach combining a microphone to capture the acoustic signals produced by tapping the fruit in the field, and a camera module for visual feature extraction and analysis.


\section{Statement of the Problem}

Accurate and objective watermelon ripeness assessment remains a persistent challenge for pre-harvest decision-making, particularly in local farming communities where ripeness assessment relies entirely on years of accumulated experience. In practice, farmers in a watermelon farm in Cagayan de Oro City harvest watermelons according to their buyers' specifications, who typically request a mixed batch — a portion already fully ripe for immediate sale and a portion harvested slightly earlier, whose perceived palatability improves over the following days due to post-harvest textural and sugar redistribution rather than true biochemical ripening. This classification requires accurately distinguishing between multiple ripeness stages simultaneously, a skill that only experienced farmers reliably possess. While experienced farmers supplement this assessment with week-based tracking of growth duration, ripeness progression varies with seasonal conditions such as weather, soil quality, and irrigation, meaning that harvest timing cannot be reliably determined by growth duration alone without physical assessment of individual fruit.

For new and incoming farmers, replicating this classification without a standardized reference is difficult, as the traditional thumping technique and visual inspection of the rind provide no measurable output and cannot be easily learned or verified without extensive field experience. This knowledge gap creates inconsistency in harvest timing decisions and leaves new farmers without a practical tool to guide pre-harvest assessment.

Efforts to address this challenge have led researchers to explore nondestructive testing techniques as objective alternatives to manual assessment. Acoustic-based approaches have demonstrated that tapping-induced sound signals carry meaningful information about internal fruit quality, with frequency and magnitude patterns showing strong correlation with soluble solids content and maturity stage, yet these methods have consistently been validated only under controlled laboratory conditions rather than real open-field environments \citep{okafor_2020_design, pamungkas_2021_evaluation, wongkalasin_2021_watermelon}. Their performance in open field environments remains unreliable, as environmental noise from wind, machinery, and handling interference significantly degrades signal quality \citep{pamungkas_2021_evaluation}. 

Computer vision-based approaches, on the other hand, are either restricted to surface-level external indicators and cannot assess internal quality attributes such as flesh firmness or sugar content \citep{defang_2024_intelligent, pamungkas_2021_evaluation}, or achieve internal quality estimation only through expensive stationary hyperspectral equipment incompatible with portable farm-level deployment \cite{shah_2025_a, qi_2024_application}. Despite these efforts, no portable integrated system combining both acoustic and visual modalities through a unified machine learning classifier for real-time field use has been identified in the literature.

This study addresses this gap by developing a portable device that analyzes the acoustic response and rind visual characteristics of watermelons, integrating both inputs through a machine learning classifier to enable objective and real-time ripeness classification directly in the field and providing local farmers, particularly those new to the practice,  with a reliable and standardized alternative to experience-based manual assessment.
 

\section{Objectives of the Study}

This study aims to develop a portable, non-destructive watermelon ripeness detection system that integrates acoustic signal processing and machine learning-based analysis for real-time, on-site classification. Specifically, the study aims to accomplish the following objectives:

\begin{enumerate}
	\item To design an acoustic–optical system that uses digital signal processing to extract resonant frequencies from vibration signals and a camera to record external watermelon features.
	\item To develop a portable, field-deployable system for real-time classification, and assess its usability.
	\item To validate system accuracy and reliability against manual and destructive reference methods using repeated-trial consistency and metrics. 

\end{enumerate}

\section{Significance of the Study}

The development of a portable, non-destructive watermelon ripeness assessment system contributes to the broader goals of sustainable agricultural development, particularly the United Nations Sustainable Development Goals of Zero Hunger (SDG 2), Responsible Consumption and Production (SDG 12), and Partnerships for the Goals (SDG 17). By improving harvest timing accuracy and reducing post-harvest losses, the system directly supports food security and the promotion of sustainable agriculture aligned with SDG 2. Its capacity to minimize produce waste along the supply chain addresses SDG 12, which calls for more efficient and responsible use of resources in food production and distribution. Furthermore, the methodologies and findings of this research contribute to the body of knowledge available to the wider scientific and agricultural community, supporting the technology transfer and knowledge-sharing objectives of SDG 17. The findings and output of this research will be significantly beneficial to the following:
 
\textbf{Local Farmers and Agricultural Producers.} This study provides an objective tool that replaces subjective manual ripeness assessment methods. By providing accurate harvest timing, farmers in regions such as Northern Mindanao can minimize post-harvest losses, ensure consistent fruit quality, and command better market prices in urban centers like Cagayan de Oro City.

\textbf{Classifiers.} The proposed system aims to assist classifiers in making faster and more consistent ripeness assessments, reducing the time spent on manual inspection per unit. This allows farm gate operations to process watermelons more efficiently before loading them onto trucks for distribution to other parts of the country.

\textbf{Consumers and Retailers.} The system will help ensure that only ripe watermelons will reach the market. This increases consumer confidence and satisfaction by guaranteeing internal quality attributes—such as sweetness and firmness—reducing the financial risk of purchasing over-ripe or hollow-heart produce.

\textbf{Future Researchers.} The data and methodologies documented in this study may inform further research into other fruit varieties or the integration of more advanced artificial intelligence models for real-time agricultural diagnostics in uncontrolled environments.

\section{Scope and Limitations}

The proposed system is designed to function as a portable, handheld tool for pre-harvest watermelon ripeness assessment. It integrates acoustic signal processing and image-based analysis to evaluate both internal and external indicators of maturity, classifying watermelons into two categories: ripe and unripe. The local image dataset will be sourced from two locations, ripe watermelon rind images will be collected from local public markets where watermelons have already been confirmed ready for sale, while unripe samples will be captured directly from pre-harvest fruit still attached to the vine at the Brgy. Balubal farm in Cagayan de Oro City, Northern Mindanao. Data collection will be conducted across multiple visits on different days to account for natural variation in field and market conditions. To supplement the local dataset, publicly available online image datasets will also be incorporated into the training pipeline for the image classification model. Ground truth ripeness labels will be established through physical cutting and inspection of each sampled watermelon to verify its internal condition. All system development and testing will be conducted under natural field and market conditions to reflect actual farm use.

The system focuses primarily on the Sweet 16 watermelon variety, which is the principal cultivar grown at the Brgy. Balubal farm. The system evaluates one watermelon at a time and is intended exclusively for pre-harvest assessment of fruit still attached to the vine. It does not support simultaneous or large-scale batch assessments, and post-harvest quality, storage behavior, and shelf life fall outside its scope. 

The system is designed to operate under natural outdoor field conditions, accounting for typical variations in ambient lighting throughout the day. The image-based classification component is optimized for outdoor rind capture under normal daylight conditions, while the acoustic classification component is designed to tolerate moderate levels of background noise commonly present in farm environments, such as wind and ambient farm activity.

Results obtained from this study are specific to the Sweet 16 watermelon variety under the growing conditions of the Brgy. Balubal farm. As the system is developed and validated within a defined set of environmental and agricultural conditions, its classification performance may vary when applied to other watermelon varieties, farm settings, or geographic regions outside those covered in this study.

\section{Research Locale and Respondents}

The study will be conducted in the Brgy. Balubal farm in Cagayan de Oro City, Northern Mindanao. It will serve as the site for data collection and system testing. 

The respondents will be the farmers, helpers and watermelon harvesters in the Brgy. Balubal farm. 

\section{Definition of Terms}

This section provides the key terms used in the study. These are arranged according to their relevance and importance to the research.

\begin{description}
	\item[Acoustic Signal Processing.] 
	A technique used to capture and analyze the sound waves generated from a watermelon "thump" to identify frequency patterns associated with maturity. In this study, it is applied to the resonant sound produced by thumping a watermelon to extract features associated with ripeness.
	
	\item[Machine Learning.] 
   A category of computational algorithms that classify data by identifying patterns within extracted features. The proposed device employs it to determine watermelon ripeness from the processed visual inputs.

    \item[Pre-harvest Assessment.]
    The evaluation of fruit maturity while still attached to the vine to determine optimal harvest timing. The system is designed for field deployment to support this stage of the production cycle.

    \item[Non-Destructive Testing.] 
    A quality evaluation method that inspects the internal state of a subject without causing physical damage. This principle applies to the approach of the system, as ripeness is inferred entirely through acoustic and visual analysis rather than cutting or puncturing the fruit.

    \item[Non-invasive Testing.]
    A method of quality evaluation that requires direct physical breach of the interior of the subject to elicit a measurable response. The proposed system is considered non-invasive, as the thumping technique is limited to external contact and does not breach the watermelon fruit.

    \item[Thumping Technique.]
    The manual act of striking a watermelon to produce a resonant sound for analysis. The system will generate this sound using the custom thumper.
	

\end{description}

}
